---
title: "Flow-Habitat Area Curves"
author: "Department of Water Resources Healthy Rivers and Landscapes Team"
date: "`r gsub(' 0([1-9])', ' \\1', format(Sys.Date(), '%B %d, %Y'))`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

<style>
/* section header spacing */
h1, h2, h3, h4 {
  margin-top: 1.5em;
  margin-bottom: 0.5em;
}

/* figure spacing */
img {
  margin-top: 1.5em;
  margin-bottom: 0.5em;
}
</style>



# Setup

```{r set_up}
# set markdown chunk options
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(out.width = "70%")

# load ggplot2; only package I don't like calling functions from explictly  
# because plotting code gets way too crowded
library(ggplot2)

# establish colors
white <- "#ffffff"
light_grey <- "#f7f7f7"
mid_grey <- "#d3d3d3"
dark_grey <- "#898989"
black <- "#000000"

ggplot_cat_colors <- scale_colour_hue()$palette(3)
ggplot_cat_colors_light <- colorspace::lighten(ggplot_cat_colors, amount = 0.3)
ggplot_cat_colors_dark <- colorspace::darken(ggplot_cat_colors, amount = 0.3)
```



# Feather River flow data

## Data pull and cleaning

```{r create_feather_flows}
# create a function to pull and clean CDEC streamgage data
get_clean_cdec_flow <- function(station,
                                sensor,
                                dur_code,
                                start_date,
                                end_date) {
  
  flow <- CDECRetrieve::cdec_query(
    station = station,
    sensor = sensor,
    dur_code = dur_code,
    start_date = start_date,
    end_date = end_date
  )
  
  flow <- flow[!is.na(flow$datetime), ]
  flow <- flow[!is.na(flow$parameter_value), ]
  
  flow$datetime <- lubridate::as_date(flow$datetime)
  
  year_part <- lubridate::year(flow$datetime) - (lubridate::month(flow$datetime) < 10)
  flow$water_year <- paste(year_part, year_part + 1, sep = "-")
  
  flow$station_water_year <- paste(flow$location_id, flow$water_year, sep = "_")
  
  flow$water_year_start <- as.Date(paste0(year_part, "-10-01"))
  
  flow$water_year_day <- as.integer(flow$datetime - flow$water_year_start + 1)
  
  return(flow)
}

# specify start and end dates; two of the stations don't have data before 2021,
# so going with a time period where all stations have essentially complete data
start_date <- "2021-10-01"
end_date <- "2024-09-30"

# pull CDEC data for gages on the Feather River
tfb_flows <- get_clean_cdec_flow(
  station = "TFB",
  sensor = 41, # daily discharge (CFS)
  dur_code = "D", # daily timestep
  start_date = start_date,
  end_date = end_date
)

grl_flows <- get_clean_cdec_flow(
  station = "GRL",
  sensor = 41, # daily discharge (CFS)
  dur_code = "D", # daily timestep
  start_date = start_date,
  end_date = end_date
)

fsb_flows <- get_clean_cdec_flow(
  station = "FSB",
  sensor = 20, # hourly discharge (CFS) - daily isn't available
  dur_code = "H", # hourly timestep
  start_date = start_date,
  end_date = end_date
)

# convert FSB hourly data to daily means
fsb_flows <- fsb_flows |>
  dplyr::group_by(datetime) |>
  dplyr::mutate(parameter_value = mean(parameter_value, na.rm = TRUE)) |>
  dplyr::ungroup() |>
  unique()

# create a single flows object
flows <- rbind(tfb_flows, grl_flows, fsb_flows)

# add an exceedance attribute
flows <- flows |>
  dplyr::group_by(location_id) |>
  dplyr::mutate(exceedance = 1 - (rank(parameter_value, ties.method = "average") / (dplyr::n() + 1))) |>
  dplyr::ungroup()
```



## Visualizations

```{r view_feather_hydrographs, echo = FALSE, out.width = "100%"}
# create a factor to intelligibly label facets
flows$facet_label <- factor(
  flows$location_id,
  levels = c("TFB", "GRL", "FSB"),
  labels = c("Thermalito Fish Barrier", "Gridley", "Star Bend"),
)

# plot hydrographs
ggplot(
  data = flows,
  aes(
    x = water_year_day,
    y = parameter_value,
    color = as.factor(water_year),
    group = water_year
  )) +
  annotate(
    "rect",
    xmin = 93, xmax = 274,
    ymin = -Inf, ymax = Inf,
    fill = mid_grey,
    alpha = 0.5
  ) +
  geom_line() +
  facet_wrap(~ facet_label) +
  scale_x_continuous(
    breaks = c(1, 32, 62, 93, 124, 152, 183, 213, 244, 274, 305, 336),
    labels = c("Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    breaks = seq(from = 0, to = 50000, by = 5000),
    expand = c(0, 0)
  ) +
  labs(
    y = "dischage (cfs)",
    color = "water year"
  ) +
  guides(color = guide_legend(label.hjust = 0.5)) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    axis.line = element_line(color = dark_grey),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

```{r view_feather_boxplots, echo = FALSE}
# plot boxplots of discharge at each site
ggplot(data = flows, aes(x = facet_label, y = parameter_value)) +
  geom_boxplot() +
  scale_y_continuous(
    labels = scales::comma,
    breaks = seq(from = 0, to = 50000, by = 5000)) +
  labs(
    x = NULL,
    y = "discharge (cfs)"
  ) +
  theme_minimal() +
  theme(
    axis.line.x = element_line(color = dark_grey),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```

```{r view_feather_exceedance}
# plot exceedance probabilities; data are log-transformed for better 
# visualization but y-axis labels are back-transformed for easy interpretation
ggplot(
    data = flows,
    aes(x = exceedance, y = parameter_value, color = facet_label)
  ) +
  geom_line() +
  scale_x_continuous(
    labels = scales::percent,
    breaks = seq(0, 1, by = 0.1),
    limits = c(0, 1)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    trans = "log10",
    breaks = c(1000, 2000, 5000, 10000, 20000, 50000)
  ) +
  labs(
    x = "percent exceedance",
    y = "discharge (cfs)",
    color = "location"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = dark_grey),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



# Feather River flow-habitat curves

## Data pull

```{r import_sbrs_from_edi}
package_id <- "edi.1511.1"

entity_names <- EDIutils::read_data_entity_names(packageId = package_id)
entity_names$entityName <- snakecase::to_any_case(
  entity_names$entityName,
  case = "snake"
)

for (i in seq_len(nrow(entity_names))) {
  entity_id <- entity_names$entityId[i]
  entity_name <- entity_names$entityName[i]
  
  data_entity <- EDIutils::read_data_entity(
    packageId = package_id,
    entityId = entity_id
  )
  
  data_read <- readr::read_csv(file = data_entity)
  
  assign(x = entity_name, value = data_read)
}

feather_hab_flow_addl <- dplyr::filter(
  habitat_flow_relationships,
  watershed == "Feather River",
  scenario == "VA additional acres",
  habitat_type == "Instream rearing"
)

feather_hab_flow_addl <- dplyr::select(
  feather_hab_flow_addl,
  flow_cfs,
  habitat_area_acres
)
```



## Interpolation set up

### Parameters

```{r specify_interp_parameters}
# specify a margin of error; selecting 5% as prescribed by draft POI
moe_prop <- 0.05
```

### Functions

```{r specify_interp_functions}
# create a function that interpolates between two coordinate pairs with a sine curve
interpolate_sine <- function(x_start, x_end, y_start, y_end, n) {
  # generate n equally spaced x values between x_start and x_end
  x_values <- seq(x_start, x_end, length.out = n)
  
  # calculate the amplitude and vertical shift based on y_start and y_end
  amplitude <- (y_end - y_start) / 2
  y_center <- (y_start + y_end) / 2
  
  # scale x values to cover a half sine wave, ensuring it goes from -1 to 1
  y_values <- y_center + amplitude * sin((x_values - x_start) * (pi / (x_end - x_start)) - (pi / 2))
  
  # return a dataframe of x and y coordinates
  return(data.frame(x = x_values, y = y_values))
}

# create a function that interpolates between two coordinate pairs with an ellipse arc
interpolate_ellipse <- function(x_start,
                                x_end, y_start,
                                y_end, n,
                                quadrant = "ur") {
  # set theta range and center based on the selected quadrant (upper-right,
  # upper-left, bottom-right, or bottom-left)
  if (quadrant == "ur") {
    theta <- seq(0, pi / 2, length.out = n)
    x_center <- min(x_start, x_end)
    y_center <- min(y_start, y_end)
  } else if (quadrant == "ul") {
    theta <- seq(pi / 2, pi, length.out = n)
    x_center <- max(x_start, x_end)
    y_center <- min(y_start, y_end)
  } else if (quadrant == "br") {
    theta <- seq(-pi / 2, 0, length.out = n)
    x_center <- min(x_start, x_end)
    y_center <- max(y_start, y_end)
  } else if (quadrant == "bl") {
    theta <- seq(pi, 3 * pi / 2, length.out = n)
    x_center <- max(x_start, x_end)
    y_center <- max(y_start, y_end)
  } else {
    stop("invalid quadrant, choose from 'ul', 'ur', 'br', or 'bl'")
  }
  
  # calculate the semi-major and semi-minor axes
  semi_major_axis <- abs(x_end - x_start)
  semi_minor_axis <- abs(y_end - y_start)
  
  # calculate x and y coordinates along the elliptical arc
  x_values <- x_center + semi_major_axis * cos(theta)
  y_values <- y_center + semi_minor_axis * sin(theta)
  
  # return a dataframe of x and y coordinates
  return(data.frame(x = x_values, y = y_values))
}
```



## Linear interpolation

```{r create_linear_interp}
# add margin of error columns
feather_hab_flow_addl <- dplyr::mutate(
  feather_hab_flow_addl,
  habitat_area_acres_lower = (1 - moe_prop) * habitat_area_acres,
  habitat_area_acres_upper = (1 + moe_prop) * habitat_area_acres
)

# visualize the linear interpolation with a 5% margin of error
ggplot(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres)
  ) +
  geom_ribbon(
    aes(ymin = habitat_area_acres_lower, ymax = habitat_area_acres_upper),
    fill = mid_grey,
    alpha = 0.5
  ) +
  geom_line(color = black, size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  scale_y_continuous(breaks = seq(from = 2, to = 6, by = 0.5)) +
  labs(
    title = "Linear interpolation",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



## Cubic spline interpolation (unanchored)

I don't think this is particularly helpful but still making it as a wild edge
case since this is mathematically a defensible mechanism for interpolation.

```{r create_cubic_spline_interp}
# create the cubic spline curve values
cubic_spline <- as.data.frame(stats::spline(
  x = feather_hab_flow_addl$flow_cfs,
  y = feather_hab_flow_addl$habitat_area_acres,
  n = 1000
))

colnames(cubic_spline) <- c("flow_cfs", "habitat_area_acres")

cubic_spline <- dplyr::mutate(
  cubic_spline,
  habitat_area_acres_lower = habitat_area_acres * (1 - moe_prop),
  habitat_area_acres_upper = habitat_area_acres * (1 + moe_prop)
)

# visualize the cubic spline interpolation with a 5% margin of error
ggplot() +
  geom_ribbon(
    data = cubic_spline,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = mid_grey,
    alpha = 0.5
  ) +
  geom_line(
    data = cubic_spline,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = black,
    size = 1
  ) +
  geom_point(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    size = 2
  ) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  labs(
    title = "Cubic spline interpolation (unanchored)",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



## Cubic spline interpolation (anchored)

```{r create_cubic_spline_interp_anchored}
# add slightly arbitrary anchor points between the second pair of coordinates
anchor_rows <- tibble::tribble(
  ~flow_cfs, ~habitat_area_acres, ~habitat_area_acres_lower, ~habitat_area_acres_upper,
  800,       4.9,                 5 * (1 - moe_prop),        5 * (1 + moe_prop),
  1030,      5.0,                 5 * (1 - moe_prop),        5 * (1 + moe_prop),
  1400,      4.75,                5 * (1 - moe_prop),        5 * (1 + moe_prop)
)

feather_hab_flow_addl_anchored <- dplyr::bind_rows(
  feather_hab_flow_addl,
  anchor_rows
)

# create the anchored ubic spline curve values
cubic_spline_anchored <- as.data.frame(stats::spline(
  x = feather_hab_flow_addl_anchored$flow_cfs,
  y = feather_hab_flow_addl_anchored$habitat_area_acres,
  n = 1000
))

colnames(cubic_spline_anchored) <- c("flow_cfs", "habitat_area_acres")

cubic_spline_anchored <- dplyr::mutate(
  cubic_spline_anchored,
  habitat_area_acres_lower = habitat_area_acres * (1 - moe_prop),
  habitat_area_acres_upper = habitat_area_acres * (1 + moe_prop)
)

# visualize the cubic spline interpolation with a 5% margin of error
ggplot() +
  geom_ribbon(
    data = cubic_spline_anchored,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[1],
    alpha = 0.5
  ) +
  geom_line(
    data = cubic_spline_anchored,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[1],
    size = 1
  ) +
  geom_point(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    size = 2
  ) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  labs(
    title = "Cubic spline interpolation (anchored)",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



# Sine interpolation

```{r create_sine_interp}
# create a half-cycle sine interpolation between first two coordinate pairs
first_sine_interp <- interpolate_sine(
  x_start = as.numeric(feather_hab_flow_addl[1, "flow_cfs"]),
  x_end   = as.numeric(feather_hab_flow_addl[2, "flow_cfs"]),
  y_start = as.numeric(feather_hab_flow_addl[1, "habitat_area_acres"]),
  y_end   = as.numeric(feather_hab_flow_addl[2, "habitat_area_acres"]), 
  n = 100
)

# create a half-cycle sine interpolation between second two coordinate pairs
second_sine_interp <- interpolate_sine(
  x_start = as.numeric(feather_hab_flow_addl[2, "flow_cfs"]),
  x_end   = as.numeric(feather_hab_flow_addl[3, "flow_cfs"]),
  y_start = as.numeric(feather_hab_flow_addl[2, "habitat_area_acres"]),
  y_end   = as.numeric(feather_hab_flow_addl[3, "habitat_area_acres"]), 
  n = 100
)

# combine into a single sine interpolation
sine_interp <- rbind(first_sine_interp, second_sine_interp)
colnames(sine_interp) <- c("flow_cfs", "habitat_area_acres")

sine_interp <- dplyr::mutate(
  sine_interp,
  habitat_area_acres_lower = habitat_area_acres * (1 - moe_prop),
  habitat_area_acres_upper = habitat_area_acres * (1 + moe_prop)
)

# visualize the half-cycle sine interpolation with a 5% margin of error
ggplot() +
  geom_ribbon(
    data = sine_interp,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[2],
    alpha = 0.5
  ) +
  geom_line(
    data = sine_interp,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[2],
    size = 1
  ) +
  geom_point(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    size = 2
  ) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  labs(
    title = "Half-cycle sine interpolation",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



# Ellipse interpolation

```{r}
# create an elliptical arc between the first two coordinate pairs
first_ellipse_interp <- interpolate_ellipse(
  x_start = as.numeric(feather_hab_flow_addl[1, "flow_cfs"]),
  x_end   = as.numeric(feather_hab_flow_addl[2, "flow_cfs"]),
  y_start = as.numeric(feather_hab_flow_addl[1, "habitat_area_acres"]),
  y_end   = as.numeric(feather_hab_flow_addl[2, "habitat_area_acres"]), 
  n = 100,
  quadrant = "ul"
)

# create an elliptical arc between the second two coordinate pairs
second_ellipse_interp <- interpolate_ellipse(
  x_start = as.numeric(feather_hab_flow_addl[2, "flow_cfs"]),
  x_end   = as.numeric(feather_hab_flow_addl[3, "flow_cfs"]),
  y_start = as.numeric(feather_hab_flow_addl[2, "habitat_area_acres"]),
  y_end   = as.numeric(feather_hab_flow_addl[3, "habitat_area_acres"]), 
  n = 100,
  quadrant = "ur"
)

# combine into a single elliptical arc interpolation
ellipse_interp <- rbind(first_ellipse_interp, second_ellipse_interp)
colnames(ellipse_interp) <- c("flow_cfs", "habitat_area_acres")

ellipse_interp <- dplyr::mutate(
  ellipse_interp,
  habitat_area_acres_lower = habitat_area_acres * (1 - moe_prop),
  habitat_area_acres_upper = habitat_area_acres * (1 + moe_prop)
)

# visualize the elliptical arc interpolation with a 5% margin of error
ggplot() +
  geom_ribbon(
    data = ellipse_interp,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[3],
    alpha = 0.5
  ) +
  geom_line(
    data = ellipse_interp,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[3],
    size = 1
  ) +
  geom_point(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    size = 2
  ) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  labs(
    title = "Ellipse interpolation",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```

```{r visualize_all_interp}
ggplot() +
  geom_ribbon(
    data = feather_hab_flow_addl,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = dark_grey,
    alpha = 0.5
  ) +
  geom_line(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = black,
    size = 1
  ) +
  geom_ribbon(
    data = cubic_spline_anchored,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[1],
    alpha = 0.5
  ) +
  geom_line(
    data = cubic_spline_anchored,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[1],
    size = 1
  ) +
  geom_ribbon(
    data = sine_interp,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[2],
    alpha = 0.5
  ) +
  geom_line(
    data = sine_interp,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[2],
    size = 1
  ) +
    geom_ribbon(
    data = ellipse_interp,
    aes(
      x = flow_cfs,
      ymin = habitat_area_acres_lower,
      ymax = habitat_area_acres_upper
    ),
    fill = ggplot_cat_colors_light[3],
    alpha = 0.5
  ) +
  geom_line(
    data = ellipse_interp,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = ggplot_cat_colors_dark[3],
    size = 1
  ) +
  geom_point(
    data = feather_hab_flow_addl,
    aes(x = flow_cfs, y = habitat_area_acres),
    color = black,
    size = 3
  ) +
  scale_x_continuous(
    limits = c(600, 3200),
    labels = scales::comma,
    breaks = seq(from = 600, to = 3200, by = 400),
    expand = c(0, 0)
  ) +
  labs(
    title = "Interpolation methods",
    x = "discharge (cfs)",
    y = "habitat area (acres)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(margin = margin(b = 20)),
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  )
```



##

```{r}
```
